---
title: "Road Network"
author: "Juan Fonseca"
format: html
---
```{r libraries}
library(tidyverse)
library(dodgr)
library(osmextract)
library(sf)
library(tmap)
library(stplanr)
```

The following code uses some functions compiled in the `MinorRoadTraffic` repository prepared by Malcolm Morgan and follows the process described in [this vignette](https://github.com/ITSLeeds/MinorRoadTraffic/blob/main/vignettes/IOW.Rmd) of the same package.

```{r minor_library}
# remotes::install_github("ITSLeeds/MinorRoadTraffic") # if not installed
library(MinorRoadTraffic)
```

## Downloading data

For this analysis, OpenStreetMap will be used for the road network.

```{r osm_get}
osm_raw <- oe_get("East Yorkshire with Hull",
                 extra_tags = c("ref", "highway", "junction", "maxspeed","oneway"))
```

An alternative source can be the [OS Open Roads](https://osdatahub.os.uk/downloads/open/OpenRoads).

## Pre-Processing OSM

### Clipping road network

The study area bounds will be used to clip the road network.
```{r load_counts_results}
sf_counts <- st_read("03_preprocessing_files/grouped_counts.geojson")
bounds <- st_read(dsn = "03_preprocessing_files/bounds.geoJSON")
```

```{r clipping}
osm_clean = extract_from_osm(osm_raw,
                             bounds,
                             col_names = c("osm_id","name","ref","highway",
                                           "junction","maxspeed","geometry","oneway"))

network = osm_clean$network
rm(osm_clean, osm_raw)
```

### Filling missing values
The `ref` attribute is filled based on the neighbouring links. 
```{r attr_filling}
network_fill = osm_fill_ref2(network)
```

## Join with other datasets

### Traffic counts
```{r load_aadf}
aadf_data <- read_csv(
  "03_preprocessing_files/aadf_data.csv",
  col_types = cols(
    group_id = col_double(),
    subgroup_id = col_double(),
    flow.2022 = col_double(),
    flow.2023 = col_double()
    )
  )
```

Joining the counts and the spatial data
```{r join_counts}
sf_aadf <- sf_counts |> 
  left_join(aadf_data,by = join_by(group_id, subgroup_id))
```

A quick visualisation
```{r map_net_counts}
tmap_mode("plot")
tm_basemap("OpenStreetMap")+
tm_shape(bounds)+
  tm_polygons(alpha = 0.3)+
  tm_shape(network)+
  tm_lines(col = "orange",alpha = 0.4)+
  tm_shape(sf_aadf)+
  tm_dots(col = "blue")
```

It is assumed that the locations of the loops are correct and the uncertainty of the coordinates is very small. 

<!-- Include something to describe the bearing check -->

The `assign_aadt_major` of the `MinorRoadsTraffic` package uses voronoi diagrams to assign the traffic flows to the major roads of the network. The assignment will be different for this analysis for the following reasons:

- The hull dataset has unidirectional traffic counts, as opposed to the bi-directional counts of the dataset by DfT . 
- Traffic counts represent the flow in a specific section of the road links, other sections before and after adjacent junctions might be different. 

A graph is created using the `dodgr` package
```{r aadf_assign}
major_ref = c("motorway",
              "motorway_link",
              "primary",
              "primary_link",
              "trunk",
              "trunk_link")

graph <- dodgr::weight_streetnet(network_fill,
                                 keep_cols = c("name","ref","highway","junction","maxspeed","oneway"),
                                 wt_profile = "motorcar"
                                 )


```

As the network has been clipped, it is possible to find some disconnected parts from the main network, these smaller component will be discarded; additionally, duplicated edges are discarded. 

```{r}
main_component <- tibble(component = graph$component) |> 
  count(component) |>
  filter(n == max(n)) |>
  pull(component)

graph_c1 <- graph[graph$component %in% main_component,] |>
  dodgr_deduplicate_graph()
```

Betweenness centrality is calculated using the contracted graph and `dodgr`
```{r centrality_calc}
graph_clean <- graph_c1 |> 
  dodgr_centrality(contract = T)
```


An `sf` object is produced from the clean network
```{r}
graph_sf <- dodgr::dodgr_to_sf(graph_clean) |>
  st_transform(27700)
```


Links within 20 metres from the count are considered and then filtered using the  stated direction; then the nearest feature/link with the correct bearing is selected. If two or more counts are assigned to the same road link/graph, the mean flow is calculated. 
```{r}
traffic_buffer <- sf_aadf |>
  st_buffer(20)

road_intersects <- st_intersects(traffic_buffer,graph_sf)

dirs_tbl <-
  tibble(
    dir_str = c(
      "southbound",
      "westbound",
      "northbound",
      "eastbound",
      "northeastbound",
      "northwestbound",
      "southeastbound",
      "southwestbound"
    ),
    t_bearing = c(180, -90, 0, 90,
                  45, -45, -135, 135)
  )

#Bearing tolerance
b_tolerance <- 90


sf_aadf$edge_id <- vapply(seq_len(nrow(traffic_buffer)),
       function(t_count){
  str_dir_bearing <- dirs_tbl$t_bearing[traffic_buffer$dir_str[t_count]==dirs_tbl$dir_str]
  
  sel_road_links = graph_sf[road_intersects[[t_count]],]
  
  if(nrow(sel_road_links)>0) {
    sel_road_links$bearing = stplanr::line_bearing(l = sel_road_links |>
                                                     st_transform(crs = 4326))
    
    sel_road_links$bearing_check = (str_dir_bearing - abs(sel_road_links$bearing))  < b_tolerance
    
    sel_road_links = sel_road_links[sel_road_links$bearing_check, ]
    
    t_edge = sel_road_links[st_nearest_feature(sf_aadf[t_count, ], sel_road_links), ] |>
      st_drop_geometry()
    
    return(t_edge$edge_id)
  } else{
    return(NA_character_)
  }
  
  },
  character(1))

edge_flow <- sf_aadf |>
  st_drop_geometry() |>
  drop_na(edge_id) |>
  summarise(across(starts_with("flow"),mean),.by = edge_id)
```

A final version of the `sf` object of the network with the known flows is produced

```{r}
graph_contr = dodgr::dodgr_contract_graph(graph_clean)

graph_sf_flows <- graph_sf |>
  left_join(edge_flow,by="edge_id")

graph_sf_flows$road_type <-
  dplyr::if_else(graph_sf_flows$highway %in% major_ref,
                 "major",
                 "minor")
```

```{r,include=FALSE,message=FALSE,warning=FALSE}
tmap_mode("plot")
```

```{r}
tm_shape(graph_sf_flows)+
  tm_lines(col = "road_type",lwd = 1,palette = "-Greens")
```

The following map shows the road network and the flows that have been assigned
```{r assign_map,message=FALSE,warning=FALSE}
tmap_mode("plot")
tm_shape(graph_sf_flows |> filter(is.na(flow.2023)))+
  tm_lines("grey",lwd = 1)+
  tm_shape(sf_aadf)+
  tm_dots("blue",alpha=0.5)+
  tm_shape(graph_sf_flows |> drop_na(flow.2023))+
  tm_lines("flow.2023",lwd = 1.5)
```

#### Junctions

Junctions are extracted from the contracted graph to avoid redundant nodes

```{r jct_ext}
junctions <- dodgr::dodgr_vertices(graph_contr) |>
  st_as_sf(coords = c("x","y"),
           crs = 4326)
```

Junctions will be classified as minor, major and minor-major depending on the adjacent links.

```{r jct_classification}
junction_class_to <- graph_sf_flows |>
  st_drop_geometry() |>
  summarise(count = n(),.by = c(to_id,road_type)) |>
  pivot_wider(names_from = road_type,values_from = count)|> 
  rename(id = to_id)

junction_class_from <- graph_sf_flows |>
  st_drop_geometry() |>
  summarise(count = n(),.by = c(from_id,road_type)) |>
  pivot_wider(names_from = road_type,values_from = count) |>
  rename(id = from_id)


junctions_classed <- junction_class_to |> 
  full_join(junction_class_from,by = "id",suffix = c(".to",".from")) |> 
  mutate(jct_type = case_when(is.na(minor.to)&is.na(minor.from)~"major",
                              is.na(major.to)&is.na(major.from)~"minor",
                              (!is.na(minor.to)&!is.na(major.from))|
                                (!is.na(minor.from)&!is.na(major.to))~"minmaj")) |> 
  select(-starts_with("m"))
rm(junction_class_from,junction_class_to)
```

Mean incoming approach flows  are calculated for all junctions, i.e. if a junction has 4 incoming roads/edges, an average of the 4 is calculated. Outgoing flows are not considered to avoid double-counting since the in and out flows should be the same.  

```{r jct_flows}
junctions_in_flows <- graph_sf_flows |>
  st_drop_geometry() |>
  summarise(across(starts_with("flow"),\(x) mean(x,na.rm = T)),.by = c(to_id))|> 
  rename(id = to_id)
```

```{r jct_consolidate}
junctions_flows <- junctions |> 
  left_join(junctions_in_flows,by = "id") |> 
  left_join(junctions_classed,by = "id")
rm(junctions_classed,junctions_in_flows,junctions)
```

The following map shows the average flow in Minor-Major junctions.

```{r minmaj_jct_plot}
tmap_mode("plot")
tm_shape(graph_sf_flows)+
  tm_lines("#87CEFA",lwd = 1)+
  tm_shape(junctions_flows |> filter(jct_type == "minmaj"))+
  tm_dots(col = "flow.2023",size = 0.2, title = "AADF 2023")
```

### LSOA data

```{r load_lsoa_data}
lsoa_data <- st_read("03_preprocessing_files/LSOA_data.gpkg")
```

Road links are joined to the LSOA based on the largest overlap. It will be assumed that the length of each road will be proportional to the population served in each LSOA, i.e. if there are two roads in an LSOA, one with 750 m and the other with 250 m, their distribution will be assumed as 75% and 25% respectively. MAUP are acknowledged to be a limitation.

```{r lsoa_to_roads}
edge_lsoa <- graph_sf_flows |>
  select(edge_id,d) |>
  st_join(lsoa_data |> select(LSOA21CD),
          largest = T) |>
  st_drop_geometry() |> 
  mutate(portion_lsoa = d/sum(d),.by = LSOA21CD) |> 
  select(-d)
```

## Network metrics

### Catchment area by roadlink

<!-- add a calculation for one-sided buffer -->

<!-- Voronoi polygons are not easy to generate for directed networks as there are overlapping links with different ids  -->

### Road density
This is an improved version of the `road_density` function in the `MinorRoadTraffic` package
```{r road_density, warning=FALSE}
zone_inter <- sf::st_intersects(lsoa_data, graph_sf_flows)

lsoa_data$road_km <- pbapply::pbvapply(seq_along(zone_inter),
                             function(zone) {
                               t_zone <- lsoa_data[zone,]
                               line <- graph_sf_flows[zone_inter[[zone]],]
                               
                               if (nrow(line) == 0) {
                                 return(0)
                               }
                               
                               line <- sf::st_intersection(t_zone, line)
                               return(sum(as.numeric(sf::st_length(line))) / 1000)
                             },
                             FUN.VALUE = numeric(1))

lsoa_data_density <-
  lsoa_data |> mutate(area_km2 = as.numeric(st_area(geom)) / 1e6,
                      road_density = road_km / area_km2)
```

A map of the road density in $km/km^2$: 

```{r map_road_density}
tm_shape(lsoa_data_density)+
  tm_polygons(col = "road_density",
              title = "road km/sq.km",
              palette = "Greens",
              border.col = "grey")
  
```


### Junction density
Similarly, the junction is calculated with the following code:
```{r jct_dens}
lsoa_junctions <- sf::st_intersects(lsoa_data,
                                    junctions_flows |> st_transform(27700))

lsoa_data_density$jct_counts <- pbapply::pbvapply(lsoa_junctions,length,integer(1))

lsoa_data_density2 <-
  lsoa_data_density |> mutate(area_km2 = as.numeric(st_area(geom)) / 1e6,
                      jct_density = jct_counts / area_km2)
```

A map with the junction density
```{r map_jct_dens}
tm_shape(lsoa_data_density2)+
  tm_polygons(col = "jct_density",
              title = "Junctions/sq.km",
              palette = c("white","blue","darkblue"),
              border.col = "grey")
```

### Centrality

The values of centrality calculated before are scaled dividing by the maximum value
```{r centrality_scaling}
graph_sf_centrality <- graph_sf_flows |>
  mutate(std.centrality = centrality/max(centrality,na.rm = T))
```


A map to explore the centrality results
```{r map_centrality}
tmap_mode("plot")
tm_shape(graph_sf_centrality)+
  tm_lines(col = "std.centrality",palette = "viridis",lwd = 1,style = "fisher")
```

### Catchment for major roads

<!-- use the dodgr isoverts functions to get different catchment areas to be used with  the one-sided buffers -->

<!-- try different isobands 5 and 25 miles were used by Selby... different scale -->


### Distance from minor-major junctions

First the mid-point of all lines are extracted 
```{r midpoint_ext}
midpoints <- dodgr::dodgr_to_sf(graph_clean) |>
  st_drop_geometry() |>
  select(edge_id) |> 
  bind_cols(
    dodgr::dodgr_to_sf(graph_clean) |> 
  st_coordinates() |> 
  data.frame() |> 
  tibble() |> 
  filter(row_number()==ceiling(n()/2),.by =L1) |>
  select(-L1)) |> 
  # Extracting ids from uncontracted graph
  left_join(graph_clean |> dodgr_vertices(),
            by = c("X"="x","Y"="y"))
```

Junctions with known flows are sub-set and then distances from all mid-points to these junctions are calculated.

```{r dist_calc}
id_jct_flows <- junctions_flows |> 
  filter(!is.na(flow.2023))

times_matrix_flows <- dodgr_times(graph_clean,
                                  from = id_jct_flows$id,
                                  to = midpoints$id,
                                  shortest = F)
```

Checking the number of unconnected links. Zero are expected as we are analysing the main component of the graph
```{r}
length(colSums(times_matrix_flows)[is.na(colSums(times_matrix_flows,na.rm = T))])
```


Using the times matrix, we extract the `id` of the nearest junction with known flows 
```{r id_nearest_jct,warning=FALSE}
fastest_all <- tibble(
  id.jct = 
    apply(times_matrix_flows, 2,
          \(x) {id_jct_flows$id[match(min(x,na.rm = TRUE), x)]}
    ),
  dist.jct =
    apply(times_matrix_flows, 2,\(x) min(x,na.rm = TRUE)
      )
  )
```

Joining the id of the nearest known count and the distance.
```{r dist_from_jct_map}
midpoints_jct <- midpoints |> select(edge_id) |> bind_cols(fastest_all)

graph_sf_centrality |>
  left_join(midpoints_jct,
            by = join_by(edge_id)) |> 
  tm_shape()+
  tm_lines("dist.jct")
```

## Consolidating all data

```{r}
sf_net_full_data <- graph_sf_centrality |>
  left_join(midpoints_jct,
            by = join_by(edge_id)) |> 
  left_join(junctions_flows |>
              st_drop_geometry() |>
              select(id,flow.2022,flow.2023),
            by = c("id.jct"="id"),
            suffix = c("",".jct")) |> 
  left_join(edge_lsoa,by = join_by(edge_id))

```


```{r,echo=FALSE}
network_data <- list(network = sf_net_full_data,
                     junctions = junctions_flows,
                     graph = graph_clean)
write_rds(network_data,file = "03_preprocessing_files/network_data.rds")
```

## Saving data
```{r save_network_data}
st_write(sf_net_full_data, "03_preprocessing_files/network_data.gpkg",
         layer = "network_data",
         delete_layer = T,
         append = T)
st_write(lsoa_data_density2,
         "03_preprocessing_files/network_data.gpkg",
         layer = "LSOA",
         delete_layer = T,
         append = T)
st_write(junctions_flows,
         "03_preprocessing_files/network_data.gpkg",
         layer = "junctions",
         delete_layer = T,
         append = T)
```




