---
title: "Road Network"
author: "Juan Fonseca"
format: html
---
```{r}
library(tidyverse)
library(dodgr)
library(osmextract)
library(sf)
library(tmap)
```

The following code uses some functions compiled in the `MinorRoadTraffic` repository prepared by Malcolm Morgan and follows the process described in [this vignette](https://github.com/ITSLeeds/MinorRoadTraffic/blob/main/vignettes/IOW.Rmd) of the same package.

```{r}
# remotes::install_github("ITSLeeds/MinorRoadTraffic") # if not installed
library(MinorRoadTraffic)
```

## Downloading data

```{r}
osm_raw <- oe_get("East Yorkshire with Hull",
                 extra_tags = c("ref", "highway", "junction", "maxspeed"))
```

## Clipping road network

The count data will be used to clip the road network.
```{r}
sf_counts <- st_read("03_preprocessing_files/grouped_counts.geojson")
```


```{r}
buffer_distance = max(st_distance(sf_counts,sf_counts))*0.15

bounds <- sf_counts |>
  make_convex_hull(dist = buffer_distance)

osm_clean = extract_from_osm(osm_raw, bounds)

network = osm_clean$network
junctions = osm_clean$junctions
rm(osm_clean, osm_raw)
```

```{r,include=FALSE}
if(!file.exists("03_preprocessing_files/bounds.geoJSON")){
  st_write(bounds,"03_preprocessing_files/bounds.geoJSON")
  }
```

## Cleaning and simplifying
The `ref` attribute is filled based on the neighbouring links. 
```{r}
network_fill2 = osm_fill_ref2(network)
```
## Saving data
```{r}
try(st_write(network_fill2,"03_preprocessing_files/network.geoJSON"),silent = T)
```
## Spatial join with counts

```{r}
aadf_data <- read_csv(
  "03_preprocessing_files/aadf_data.csv",
  col_types = cols(
    group_id = col_double(),
    subgroup_id = col_double(),
    flow.2022 = col_double(),
    flow.2023 = col_double()
    )
  )

sf_aadf <- sf_counts |> 
  left_join(aadf_data,by = join_by(group_id, subgroup_id)) |> 
  mutate(rd_name = str_trim(str_remove(desc,pattern = "^\\w*"))) |> 
  mutate(rd_name = str_remove(rd_name,pattern = "^SITE _ ")) |> 
  mutate(rd_name = str_remove(rd_name,pattern = "A63 (/ )?")) |> 
  separate_wider_delim(cols = rd_name,delim = " / ",names_sep = "rd",too_few = "debug")
```

```{r}
tmap_mode("view")
tm_basemap()+
tm_shape(bounds)+
  tm_polygons(alpha = 0.3)+
  tm_shape(network)+
  tm_lines(col = "orange",alpha = 0.4)+
  tm_shape(sf_aadf)+
  tm_dots(col = "blue")
```



